Type-Safe Modular Parsing

### 2016-12-28 paper/sections/Introduction.tex
Revising and Splitting Overview

To address the modularity limitations of Programming Languages several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice. Examples include many tools for developing Software-Product
Lines~\cite{}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such textual
composition techniques collect the code for multiple features and
merge it together when a concrete combination of features is needed
for a particular program. The typical drawback of such techniques is
the lack of modular type-checking and separate compilation:
``\emph{most feature-oriented implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''~\cite{kastner11road}\bruno{reference to ``The road to
  Feature Modularity''}. Syntactic modularization techniques have also
been applied to the problem of \emph{extensible parsing}. There are
several approaches~\cite{} that enable the development of
\emph{syntactically} modular parsers or grammars. However these
approaches do not support separate compilation or modular
type-checking either.\bruno{mention that a reason why syntactic 
modularization techniques are popular is simplicity (in implementation
and also use).}

@@

>>>>

To address the modularity limitations of Programming Languages several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice. Examples include many tools for developing Software-Product
Lines~\cite{}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such textual
composition techniques collect the code for multiple features and
merge it together when a concrete combination of features is needed
for a particular program. As Kastner~\cite{} notes, the typical
drawback of such techniques is that
``\emph{most feature-oriented implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''\bruno{reference to ``The road to
  Feature Modularity''}. Syntactic modularization techniques have also
been applied to the problem of \emph{extensible parsing}. There are
several approaches~\cite{} that enable the development of
\emph{syntactically} modular parsers or grammars. However these
approaches do not support separate compilation or modular
type-checking either.\bruno{mention that a reason why syntactic 
modularization techniques are popular is simplicity (in implementation
and also use).}

@@

\haoyuan{I suggest this part to be moved to Introduction, and only discuss why Packrat is selected among
different parser combinators in Overview.}

\begin{comment}
Although there are many parsing techniques, not all of them are
suitable for type-safe modular parsing. In particular there are many
techniques which fail to provide modular type-checking and separate
compilation. Moreover, even if modular type-checking and separate
compilation are supported, efficiency is another
concern. A parsing technique should have low overhead when applied
in a modular setting. In the remaider of this section, we eliminate
various techniques that fail to satisfy our requirements, and argue
that that Packrat parsing~\cite{Ford2002} is a suitable candidate for
type-safe modular parsing.

\paragraph{Parser Generators} The most widely use tools for parsing
are parser generators. Parser generators help users generate parsers automatically or
semi-automatically from a given grammar. There is no restriction on
the algorithm, while most of them adopts table-based LL~\cite{lewis1968syntax} and LR~\cite{knuth1965translation} parsing
algorithms.
Although efficient, the main drawback of parser generators is that they do not support
modular type-checking and separate compilation.

Modular parsing based on parser generators is supported by many libraries~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016}. Users can separate the syntax definition and related parsing code into reusable components. Then the corresponding parsers are built by their library utilities. For example, NOA~\cite{Gouseti2014} uses Java annotation processing to collect grammar information, and then generates ANTLR4 parsers. However, such generation procedure requires a whole compilation after the collection of all grammar pieces. Once the grammar changed, even slightly, grammar information must be re-collected and the parser must be re-generated. Hence those libraries only have syntactic modularity.

Generating parsers often requires full information of grammars, thus semantic modularity is difficult to achieve in this way.

\paragraph{Parser Combinators}
Comparing with the parser generators, a \textit{parser combinator}~\cite{burge1975,Wadler1985}
takes several parsers and produce a new parser as its output. Parser combinators are
popular in functional programming, where the parsers are represented
by functions and parser combinators are higher-order functions accepts
them.

At a first look, parser combinators are very suitable for our purpose, because of two
reasons. Firstly, they are naturally modular. The manner of using them
is to write small parsers and use combinators to composed them
together. The construction procedure is explicit and fully controlled
by the programmer. Secondly, each parser combinator is represented by
a piece of code, and also are the parsers it takes. Thus in a
statically typed programming language they can be statically
type-checked.
\end{comment}

### 2016-12-28 paper/sections/Overview.tex
Revising and Splitting Overview

\section{To be discussed in Introduction}

\haoyuan{I suggest this part to be moved to Introduction, and only discuss why Packrat is selected among
different parser combinators in Overview.}

Although there are many parsing techniques, not all of them are
suitable for type-safe modular parsing. In particular there are many
techniques which fail to provide modular type-checking and separate
compilation. Moreover, even if modular type-checking and separate
compilation are supported, efficiency is another
concern. A parsing technique should have low overhead when applied
in a modular setting. In the remaider of this section, we eliminate
various techniques that fail to satisfy our requirements, and argue
that that Packrat parsing~\cite{Ford2002} is a suitable candidate for
type-safe modular parsing.

\paragraph{Parser Generators} The most widely use tools for parsing
are parser generators. Parser generators help users generate parsers automatically or
semi-automatically from a given grammar. There is no restriction on
the algorithm, while most of them adopts table-based LL~\cite{lewis1968syntax} and LR~\cite{knuth1965translation} parsing
algorithms.
Although efficient, the main drawback of parser generators is that they do not support
modular type-checking and separate compilation.

Modular parsing based on parser generators is supported by many libraries~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016}. Users can separate the syntax definition and related parsing code into reusable components. Then the corresponding parsers are built by their library utilities. For example, NOA~\cite{Gouseti2014} uses Java annotation processing to collect grammar information, and then generates ANTLR4 parsers. However, such generation procedure requires a whole compilation after the collection of all grammar pieces. Once the grammar changed, even slightly, grammar information must be re-collected and the parser must be re-generated. Hence those libraries only have syntactic modularity.

Generating parsers often requires full information of grammars, thus semantic modularity is difficult to achieve in this way.

\paragraph{Parser Combinators}
Comparing with the parser generators, a \textit{parser combinator}~\cite{burge1975,Wadler1985}
takes several parsers and produce a new parser as its output. Parser combinators are
popular in functional programming, where the parsers are represented
by functions and parser combinators are higher-order functions accepts
them.

At a first look, parser combinators are very suitable for our purpose, because of two
reasons. Firstly, they are naturally modular. The manner of using them
is to write small parsers and use combinators to composed them
together. The construction procedure is explicit and fully controlled
by the programmer. Secondly, each parser combinator is represented by
a piece of code, and also are the parsers it takes. Thus in a
statically typed programming language they can be statically
type-checked.

@@

\subsection{Choosing the Parsing Technique}\label{subsec:overview-parsing}

@@

\paragraph{Left-Recursion Elimination} The top-down, recursive descent parsing strategy adopted by those parser combinator libraries cannot support left-recursive grammars directly. The common solution is to rewrite the grammar into an equivalent but not left-recursive one, so called left-recursion elimination.

A left-recursive grammar of adding integers can be rewritten as below. On the right side, we got a new grammar without left-recursion.

\begin{tabular}{m{0.4\linewidth}m{0.4\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `+' <int>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> <expr'>

<expr'> ::= <empty> \alt `+' <int> <expr'>
\end{grammar}
\end{tabular}

The main problem with left-recursion elimination is that it is a
\emph{global} transformation on a grammar. Given that the full grammar
is known, then it is possible to remove all left-recursive
cases. However, when doing modular parsing, the full grammar is not
known!

For the example above, if we extend the original grammar to support
subtraction, we must analyse the full grammar again to rewrite it.

\begin{tabular}{m{0.4\linewidth}m{0.4\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `+' <int> \alt <expr> `-' <int>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> <expr'>

<expr'> ::= <empty> \alt `+' <int> <expr'> \alt `-' <int> <expr'>
\end{grammar}
\end{tabular}

Another issue of left-recursion elimination is that it requires extra
bookkeeping work to retain the original semantics. For example, the
expression $1+2-3$ is parsed as $(1+2)-3$ in the left-recursive
grammar, but after rewrite the result is $1((+2)-3))$. The parse tree
must be transformed to recover its structure.

\huang{todo}
\bruno{This example does not appear to be problematic: adding ``-''
  does not affect the rest of the grammar. Isn't the problem that
  sometimes adding a new case to the grammar, requires changing other
parts? Otherwise I don't see why left-recursion is a problem, and we
cannot simply claim that it is a problem. You have to make a better
effort to find a problematic grammar.
}

\paragraph{Longest Match Composition}
Some parsing libraries provide combinators for alternative composition with longest match,
while others do not. Now consider the non-left-recursive grammar:
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <int> `+' <expr>
\end{grammar}
In Parsec, for instance, the parser
\begin{lstlisting}[language=PlainCode]
parser = parseInt <|> parseAdd
\end{lstlisting}
will only parse the input \inlinecode{"1 + 2"} to \inlinecode{"1"}, as \inlinecode{parseInt} successfully parses \inlinecode{"1"}
and terminates parsing. With traditional alternative
composition in recursive descent parsers, the parser in the front successfully parses a substring of the input, then it unexpectedly
finishes parsing and discards the rest, in spite that latter parsers might be able to parse the whole input.
In contrast, \inlinecode{parser = parseAdd <|> parseInt} works as expected, yet manual composition for longest match is quite
inconvenient, especially in a modular setting. When the grammar is extended with new rules, programmers are supposed to adjust
the order of parsers, which is tedious.

\paragraph{Backtracking} The need for backtracking is also problematic
in a modular setting. For example, suppose that we only have \inlinecode{import..from} statements before,
but now we want to extend the grammar with the \inlinecode{import..as} case, as shown in the third line of grammar below.

\setlength{\grammarindent}{5em}
\begin{grammar}
<stmt> ::= `import' <ident> `from' <ident>
    \alt ...
    \alt `import' <ident> `as' <ident>
\end{grammar}

Since the \inlinecode{import..from} case shares a prefix with the new case \inlinecode{import..as}, when the former case fails, we must backtrack to the beginning. Take Parsec as an example, its choice
combinator \inlinecode{<|>} only tries the second alternative if the first fails
without any token consumption. An auxiliary function \inlinecode{try} is used for explicit backtracking.

\begin{lstlisting}[language=PlainCode]
oldParser = parseImpFrom <|> parseA <|> parseB <|> ...
newParser = try parseImpFrom <|> parseA <|> parseB <|> ... <|> parseImpAs
\end{lstlisting}

Given the full grammar, we can decide where to put \inlinecode{try} for backtracking. However, with modular parsing we are unable to overlook the full grammar, hence the worst case should be considered, namely all alternatives may share common prefixes with future cases. In that case we need to backtrack for all the branches.
To avoid failures in the future, we have to add \inlinecode{try} everywhere:

\begin{lstlisting}[language=PlainCode]
parser = try parseImpFrom <|> try parseA <|> try parseB <|> ... <|> try parseImpAs
\end{lstlisting}
which results in the worst exponential time complexity.

%\huang{I've rewritten this paragraph}\bruno{The example is good, I think but the explanation is not.
%You want to say that when you know
%  the full grammar, you can figure out where ``try'' is
%  needed. Without the full grammar you'd need assume the worst
%  case. Make an effort to make your explanation cristal clear!
%I think you want to miss the abstract explanation that you give first,
%with the explanation about the concrete example.
%Start with ``The need for backtracking is also problematic
%in a modular setting. For example, suppose ...'' and synchronize
%the abstract explanation and the explanation for the example.
%}

@@

It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.

\subsection{Packrat Parsing and Our Library \name}\label{subsec:overview-library-api}

Our prototype library is implemented in Scala, with its foundation on Packrat Parsing. In the current version of Scala (2.11), Packrat Parsing is included
in the Scala standard parser combinator library, which is an external jar file. The library provides a couple of
parser combinators, together with some basic parsers. Below we present an example to illustrate Scala parsers.

\paragraph{An Example}
A Packrat parser has type \lstinline{PackratParser[E]} for some
\lstinline{E}, which indicates the type of results it produces.
%\bruno{the representation of what?}
And we use \lstinline{lexical} for its lexing.

\lstinputlisting[linerange=8-12]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=PACKRAT_EXAMPLE
\noindent
Note that the code should be defined in an enclosing type that extends
\lstinline{scala.util.parsing.combinator.syntactical.StandardTokenParsers}
for lexing and
\lstinline{scala.util.parsing.combinator.PackratParsers}.
The first two lines are used for lexing. The third line defines a
parser which can, for instance, parse \lstinline{"str(5)"} and produce
the integer \lstinline{5}, where \lstinline{numericLit} is just a basic parser
in the API. Table~\ref{tab:packrat} gives a closer look at part of the API, which covers our experiments.

\begin{table}[t]
\begin{tabular}{l}
\hline
\begin{lstlisting}
def ~[U](q: => Parser[U]): Parser[~[T, U]]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition. \\
\hline
\begin{lstlisting}
def ^^[U](f: (T) => U): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for function application. \\
\hline
\begin{lstlisting}
def ^^^[U](v: => U): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator that changes a successful result into the specified value. \\
\hline
\begin{lstlisting}
def <~[U](q: => Parser[U]): Parser[T]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition which keeps only the left result. \\
\hline
\begin{lstlisting}
def ~>[U](q: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition which keeps only the right result. \\
\hline
\begin{lstlisting}
def repsep[T](p: => Parser[T], q: => Parser[Any]): Parser[List[T]]
\end{lstlisting} \\
\hspace{.2in}- A parser generator for interleaved repetitions. \\
\hline
\begin{lstlisting}
def ident: Parser[String]
\end{lstlisting} \\
\hspace{.2in}- A parser which matches an identifier. \\
\hline
\begin{lstlisting}
def numericLit: Parser[String]
\end{lstlisting} \\
\hspace{.2in}- A parser which matches a numeric literal. \\
\hline
\begin{lstlisting}
def |[U >: T](q: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for alternative composition. \\
\hline
\begin{lstlisting}
def |||[U >: T](q0: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for alternative with longest match composition. \\
\hline \\
\end{tabular}
\caption{Part of Scala Parser API.\bruno{We should double-check that this table lists the
  combinators used in the paper.}}\label{tab:packrat}
\end{table}

\begin{table}[t]
\begin{tabular}{l}
\hline
\begin{lstlisting}
type OpenParser[Alg, R, E] = Alg => (=> R) => PackratParser[E]
\end{lstlisting} \\
\hspace{.2in}- The type constructor for modular parsers. \\
\hline
\begin{lstlisting}
def parse(p: Parser[_]): String => Unit
\end{lstlisting} \\
\hspace{.2in}- A function that applies a parser to an input, and prints the result as a string. \\
\hline
\begin{lstlisting}
implicit def <|>(x: OpenParser[Alg1, R, E], y: OpenParser[Alg2, R, E])
    : OpenParser[Alg1 with Alg2, R, E]
\end{lstlisting} \\
\hspace{.2in}- Open parser combinator for alternative with longest match composition. \\
\hline \\
\end{tabular}
\caption{Our parser combinator library \name API.\haoyuan{I don't know if I can present <|> like this.}}\label{tab:parser}
\end{table}

Finally, a generic \inlinecode{parse}
function is used for testing:

\lstinputlisting[linerange=16-19]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=PACKRAT_RUNPARSER
Now \lstinline{parse(p)("str(5)")} produces \lstinline{5} as expected.

>>>>

@@

\begin{comment}
\subsection{Choosing the Parsing Technique}\label{subsec:overview-parsing}

@@

@@

It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
